### 2011-10-18 papers/smectics/smectics.tex
2D-results update
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1465 40876441-3496-0410-a613-a2fca35b2ae2

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Large-scale Simulation of Layered Binary Fluids and Smectic Liquid Crystals in Shear Flow}}
\vspace{0.6cm}

@@

\section{Introduction}
Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurt etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid. Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibres constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character. Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in bio-sensing.

Rheology gives a powerful way to characterise macroscopic physical properties of a soft material such as its viscosity, elastic and loss moduli, which quantify, for instance, its rigidity, response to applied forces or stresses etc. While in a Newtonian, viscous fluid, shearing gives rise to a stress which is linearly proportional to the applied shear rate -- the slope being the viscosity -- in complex fluids flow is often non-Newtonian and such a simple linear relation no longer holds. Thus, viscoelastic complex fluids may be shear thickening or shear thinning, if their effective viscosity increases or decreases with applied shear. They may even spontaneously phase separate when flowing, giving rise to shear bands which are familiar in the rheology of liquid crystals in the vicinity of the isotropic-nematic transition, and in wormlike micelles.

The rheological properties of a non-Newtonian fluid are often linked to the microstructure of the material. For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate. Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals. While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1. Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics. Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via pre-shearing, or not. In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample. An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state. This leads to increased dissipation and hence to a larger viscosity. A well-known example of permeation flow is the one which occurs e.g. in cholesterics which are sheared or pushed along the direction of the cholesteric helix. However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve. For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches. Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.

Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically. This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations. At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally. In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear. For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters. In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like micelles, surfactants, smectic liquid crystals and block-copolymer.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for a system A-2D and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
The second part of our discussion consists of simulations in three dimensions using identical parameters to system A-2D and a box size $N_x\times N_y\times N_z=128\times256\times128$.
We will refer to this system as system A-3D.
Table \ref{tab1} provides an overview of all simulation parameters.

\begin{table}[htp!]
\small
\caption{Simulation parameters: The systems were first equilibrated from spinodal decomposition at zero shear rate (EQ), followed by various shear rates, ranging from $\dot{\gamma}=2\cdot10^{-4}$ till $1.25\cdot10^{-5}$. For some runs a final switch off (SO) at zero shear rate was performed. The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
  \label{tab1}
%  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lllll}
  \begin{tabular*}{0.495\textwidth}{lllll}
\hline
run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
\hline
system A & & & &\\
\hline 
703 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
704 & $1024\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
705 & $1024\times512\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
\hline
767 & $512\times256\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
775 & $512\times256\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
769 & $512\times256\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
771 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
772 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
774 & $512\times256\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
\hline
792 & $256\times128\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
\hline
786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
\hline
system B & & & &\\
\hline
707 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
710 & $1024\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
765 & $1024\times512\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
\hline
776 & $512\times256\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
777 & $512\times256\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
778 & $512\times256\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
780 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
781 & $512\times256\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
782 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
\hline
system C & & & &\\
\hline
998 & $256\times128\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
1001 & $256\times128\times1$ & 320-640& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
1002 & $128\times256\times128$ & 320-640 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
\hline
\end{tabular*}
\end{table}

\subsection*{Results in Two Dimensions}
There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}. 
We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom). Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
\label{fig1}
\end{figure}

This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A-2D with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
If shear flow is imposed, the lamellae start to align with the flow.

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run707_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run710_800.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
\label{fig2}
\end{figure}

In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing deviates from its equilibrium value.
These regions are advected with the flow while they undergo an affine stretching.
The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
However, the reported results differ from our observations in the following aspects.
First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in band-like patterns that are wrapped around the periodic boundaries.
Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
These differences may be a direct consequence of the different boundary conditions.
While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.

\begin{figure}[htp!]
\centering
\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
\label{fig3}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right). The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
\label{fig4}
\end{figure}

To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\ 
After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
Fig. \ref{fig2} shows equivalent states for the softer system B.
The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
The lamellar alignment in flow direction after switch off is more pronounced.\\ 
Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
Where the system is extended and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
Large positive values occur in the regions where the lamellae are squeezed.
The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
Despite the local negativity of quantities, the overall average are still positive for all times.
Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
\vspace{0.6cm}

@@

\section{Introduction}

Instances of soft complex fluids are abound in our everyday world. 
Examples include colloidal suspensions, 
where particles or droplets are suspended in a viscous
fluid, living tissues, and liquid crystalline fluids. Well known
examples of colloidal suspensions include paint, cosmetics, and
food products such as ketchup, mayonnaise, and yogurt.
The intracellular network of actin fibres constituting the
cytoskeleton of mammalian cells is a polymeric fluid.
The same is true of the DNA making up the bacterial nucleoid and
eukaryotic chromosomes, and the viral genome of bacteriophages,
where enhanced confinement imparts a strong liquid crystalline
character to the system.
Liquid crystalline fluids themselves, with or without colloidal
inclusions, are of of wide technological importance in devices
such as electronic displays and sensors.

The behaviour under flow, or rheology, of such materials is important
to understand, as it affects many industrial and manufacturing processes.
At the same time, rheological measurements provide a powerful way to
characterise macroscopic properties of a soft material.
The response to applied forces or stresses can be used to quantify,
e.g., the viscosity, or elastic and loss moduli of soft materials.
On shearing, a simple Newtonian fluid exhibits a stress which is
linearly proportional to the applied shear rate; the
constant of proportionality is just the viscosity. In marked
contrast, complex fluids flow is often non-Newtonian and such a simple
linear relation no longer holds. On shearing, the effective viscosity
of a complex viscoelastic fluid may increase or decrease, corresponding
to shear thinkening or shear thinning behaviour. Complex fluids may even
spontaneously phase separate when flowing, giving rise to shear bands.
These are familiar in the rheology of liquid crystals in the vicinity of
the isotropic-nematic transition, and in wormlike micelles.

The rheological properties of a non-Newtonian fluid are typically linked
to the microstructure of the material. Particularly relevant to our
current work is the rheology of layered systems, such as smectics and
cholesterics. Here, experiments may show different rheological results
depending on whether the system has been ordered prior to measurement
(for example, by pre-shearing).
For instance, in the flow of polydomain liquid crystals, which are not
uniformly ordered, the existence of regions with conflicting director
patterns leads to a non-linear relation between shear stress and
applied shear rate.
Another example of strongly non-Newtonian rheology is region I flow
in polymeric nematic liquid crystals. While our understanding of the
physics of region I flow is far from complete, it is thought that a
defect network may form and lead to much enhanced viscosity, with
marked shear thinning. In both these cases, stiffening is usually
associated to the presence of defects in the sample.

Another intriguing type of flow which is possible in layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
defect pattern intact in steady state. This leads to increased dissipation
and hence to a larger viscosity. A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix. However, one would also expect permeation flows
to be possible when a non-translationally invariant defect pattern is
present in the system; this might be the case in quenched smectics, for
instance. In addition to their non-linear shear-stress curve, lamellar
systems are also of interest for a number of other phenomena.
In block-copolymers one may observe, under some conditions, shear-induced
phase transitions from the microphase separated state to ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
desirable to extend our theoretical understanding of such switches.
Further, lyotropic layered liquid crystals have sometimes been observed
to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
shear \cite{Diat93}. The mechanism underlying such pattern formation has
remained elusive to date.

Despite a wealth of experimental results, and the importance of rheology
to many commercial applications, theoretical progress in understanding
flow behaviour of complex fluids has remained a challenge. This is largely
because of the non-linear nature of the equations of motion, and the
coupling of the flow to different microstructures. These either require
drastic approximations or, for practical purposes, resort to computer
simulation. Even here, realistic 3-dimensional calculations are
computationally expensive.
In this work, our focus is the physics of smectics under shear.
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Further, a number of these studies have used solid walls
to drive shear in the hydrodymamics and, inconsistently, sliding periodic
boundaries for the composition variable. Such simulations have a number of
drawbacks: solid walls to drive shear limits the attanable shear rate in
larger systems and so exacerbates finite size effects; mixed boundary
conditions make it difficult to disentangle bulk and boundary effects
in the rheology. In this work, we try to overcome all of these issues.
(One notable example which does tackle a large 3-dimensional simulation
of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)

We describe smectics via a Brazovskii free energy functional
\cite{Brazovskii75}, which provides a robust framework to study systems
like micelles, surfactants, smectic liquid crystals and block-copolymers.
The governing equations in the model are a convection-diffusion equation
for the scalar order parameter, which represents the density difference
of two fluid components, and the Navier-Stokes equation describing
hydrodynamics. A stress tensor that contains extra contributions owing
to the presence of interfaces between the components is included.
These dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
We use consistent sliding periodic (Lees-Edwards) boundary conditions for
both convection-diffusion and Navier-Stokes equations. This allows us to
probe the true bulk rheology of the smectic phase for large systems and
high shear rates, as quantified by an appropriate Peclet number.

% KS: this can be moved to existing appendix
%The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
%The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
The system size was $N_x\times N_y\times N_z=512\times 1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
The distance from the critical point was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq. \ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
By this we obtained a slightly different stress response with the system with lower order appearing somwhat sofer.
The lamellar distance was identical in both cases.\\
The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
Table \ref{tab1} gives an overview of all simulation parameters.

\begin{table}[htp]
\small
\caption{Simulation parameters: Starting from spinodal decomposition the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\cdot10^{-5}$ to $2\cdot10^{-4}$. In some cases we performed a final switch off run (SO) at zero shear rate. The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
\label{tab1}
\begin{tabular*}{0.495\textwidth}{lllll}
\hline
run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
\hline
system A & & & &\\
\hline 
703 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
704 & $512\times1024\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
705 & $512\times1024\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
\hline
767 & $256\times512\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
775 & $256\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
769 & $256\times512\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
771 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
772 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
774 & $256\times512\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
\hline
792 & $128\times256\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
\hline
786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
\hline
system B & & & &\\
\hline
707 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
710 & $512\times1024\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
765 & $512\times1024\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
\hline
776 & $256\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
777 & $256\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
778 & $256\times512\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
780 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
781 & $256\times512\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
782 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
\hline
system C & & & &\\
\hline
998 & $128\times256\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
1001 & $128\times256\times1$ & 320-960& $1\cdot10^{-4}$ & $2\cdot10^{-4}$ \\
1000 & $128\times256\times1$ & 320-1280& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
1003 & $128\times256\times128$ & 320-960 & $1\cdot10^{-4}$ & $2\cdot10^{-4}$\\
1002 & $128\times256\times128$ & 320-1280 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
\hline
\end{tabular*}
\end{table}

\subsection*{Results in Two Dimensions}

Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq. \ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
In previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems B has a lower average order parameter and is closer to the critical point.
Higher viscosities lead to shorter interfaces, whereas lower values typically yield longer lamellae.
This effect has been reported before \cite{Gonnella97} and turned out to be more pronounced for the softer system B-2D, although not as quite to the same extent as in the former case.
In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom). Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
\label{fig1}
\end{figure}

The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
After a short transient phase during which the order parameter remains rather homogenous, expanded and compressed regions emerge with a lamellar spacing that deviates from the equilibrium value.
This is depicted in the centre picture of Fig. \ref{fig1} for shear rate $\gd= 1\e{-4}$.
The inhomogeneities remain present in the steady state, are advected by the flow and undergo continuously affine transformations.
A buckling instability above a critical shear rate has been described earlier by Gonnella et al. \cite{Gonnella98} and Kumaran et al. \cite{Kumaran2001}.
It was suggested that the density of defects defects get introduced into a lamellar liquid crystal due to dilation and annealing occurs due to application of shear.
This would lead to irregularities in the sample a dynamical process where the steady state defect density is determined by balance between the rate of introduction and rate of annealing of defects.
In a more recent paper Kumaran et al. state that they observe similar inhomogeneities only above a certain system size \cite{Kumaran2011}, which agrees well with our observation.
We found that shear alignment is more likely to take place in smaller systems of $N_x\times N_y\times N_z=128\times 256\times 1$.
However, this trend is not necessarily present at higher shear rates, a fact that we will discuss later pertaining to our simulations in three dimensions.\\
Kumaran et al. applied only one shear rate and focussed on the dependence of shear alignment on the Schmidt number $Sc$, defined as the ratio between kinematic viscosity and diffusivity.
Their values for $Sc$ are of the order ${\cal O}(1)$ (ours are roughly ${\cal O}(10^2-10^3)$), which is surprisingly low given the fact that the appearance of their lamellae does not seem to be dominated by heavy fluctuations.
However, identical to our results they observe enhanced shear alignement at lower Schmidt numbers, i.e. higher diffusivities.\\
Kumaran et al. also identified a second mechanism of defect creation that is due to layer compression and splitting.
Although we have not checked this in detail, this is also very likely the case in our results as the lamellar pattern is strongly compressed and heterogenous as shown in Fig. \ref{fig1}.\\
After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
Fig. \ref{fig2} shows equivalent states for system B-2D, which is closer to the critical point.
All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
Another shear-induced structure in binary lamellar systems was reported in \cite{Xu06b}.
The reported results differ from our observations in some aspects.
Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
These differences may be simply a direct consequence of the different boundary conditions that were applied.
While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.\\
To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run707_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run710_800.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
\label{fig2}
\end{figure}
